---
title: "Visualization"
output: html_document
---

Explore https://ggplot2.tidyverse.org/ and try what it says under Usage.

```{r}
library(ggplot2)

ggplot(mpg, aes(displ, hwy, colour = class)) +
  geom_point()
```

## Setup

The setup chunk always runs before anything else.

```{r setup}
# install.packages("tidyverse")
library(tidyverse)
```

## Data

Copy-paste.

```{r}
# National Parks in California
ca <- read_csv("https://bit.ly/03-data-ca")
# Acadia National Park
acadia <- read_csv("https://bit.ly/03-data-acadia")
# Southeast US National Parks
se <- read_csv("https://bit.ly/03-data-se")
# 2016 Visitation for all Pacific West National Parks
visit_16 <- read_csv("https://bit.ly/03-data-visit-16")
# All Nationally designated sites in Massachusetts
mass <- read_csv("https://bit.ly/03-data-mass")
```

Preview the data with `View()` and `head()`.

```{r}
head(ca)
```

## Build

* Adapt the first example. Plot the data ca, with year on x and visitors on y.
* What relationship do you expect to see between year and the number of visitors?

```{r}
ggplot(data = ca, aes(x = year, y = visitors)) +
  geom_point()
```

* Now map the values of the variable `park_name` to the `color` aesthetics.

```{r}
ggplot(ca, aes(year, visitors, color = park_name)) +
  geom_point()
```

## Customizing

* https://ggplot2.tidyverse.org/reference/index.html#section-scales
* https://ggplot2.tidyverse.org/reference/index.html#section-themes

```{r}
ggplot(ca, aes(year, visitors, color = park_name)) +
  geom_point() +
  labs(title = "Acadia National Park Visitation", y = "Visitation", x = "Year") +
  theme_bw()
```

## Faceting

Now using data from Southeast US National Parks.

```{r}
ggplot(se, aes(year, visitors)) +
  geom_point() +
  facet_wrap(~ state)
```

## Geoms

data = se

Why is this a poor plot? What kind of variable is `park_name`?

```{r}
ggplot(data = se, aes(x = park_name, y = visitors)) + 
  geom_point()
```

* Create a jitter plot.

```{r}
ggplot(se, aes(park_name, visitors)) + 
  geom_jitter()
```

* Create a boxplot.

```{r}
ggplot(se, aes(x = park_name, y = visitors)) + 
  geom_boxplot()
```

data = acadia

* Create a scatter plot
* Add lines
* Add a smoothed mean
* Move the smoothed mean to become the second layer. What subtle thing happened?

```{r}
ggplot(acadia, aes(year, visitors)) + 
  geom_point() +
  geom_line() +
  geom_smooth()
```

data = visit_16

* Create a bar chart. Count instances of `state`.
* Where does the variable count come from?

```{r}
ggplot(visit_16, aes(state)) + 
  geom_bar()
```

* Create a barchart of the number of visitors per state.
* Map the bar fill (colour) to `park_name`.
* Dodge the bars to allow side by side comparison.

```{r}
ggplot(visit_16, aes(state, visitors, fill = park_name)) + 
  geom_col(position = "dodge")
```




## Leftovers

## Set vs map

* Set the colour of all points to blue.

```{r}
ggplot(data = ca) +
  geom_point(mapping = aes(x = year, y = visitors), color = "blue")
```

- What is wrong in this code?

```{r}
ggplot(data = ca) +
  geom_point(mapping = aes(x = year, y = visitors, color = "blue"))
```

## Facet

* Add `facet_wrap(~ park_name)` to facet (split) the data by `park_name`. The plot is now clear evern without colour.

(I'll ommit the name of the first arguments. They are optional.)

```{r}
ggplot(ca) +
  geom_point(aes(x = year, y = visitors)) +
  facet_wrap(~ park_name)
```

## Geoms


* Instead of representing you data with points, we now use a smooth mean of the data.

```{r}
ggplot(ca) +
  geom_smooth(aes(x = year, y = visitors)) 
```

(Smoothers fit a model to your data and then plot predictions from the model.)

- Use `geom_line()` and `color = park_name` to compare each park against the overall trend.

```{r}
ggplot(ca) +
  geom_line(aes(x = year, y = visitors, color = park_name)) +
  geom_smooth(aes(x = year, y = visitors)) 
```

* Use `geom_bar()` to create a bar chart  FIXME

```{r}
ggplot(ca) +
  geom_bar(aes(x = park_name))
```

(Bar charts, histograms, and frequency polygons bin your data and then plot bin counts, the number of points that fall in each bin.)

## Arranging and exporting plots





***

# Take aways

You can use this code template to make thousands of graphs with **ggplot2**.

```{r eval = FALSE}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```
