---
title: "Visualization"
output: html_document
---

## Setup

The first chunk in an R Notebook is usually titled "setup," and by convention includes the R packages you want to load. Remember, in order to use an R package you have to run some `library()` code every session. Execute these lines of code to load the packages. 

## Your turn 1

```{r setup}
# install.packages("tidyverse")
library(tidyverse)
```

## Datasets

## Your turn 1

Type and run this code to load one dataset. Pay strict attention to spelling, capitalization, and parentheses!

```{r}
# National Parks in California
ca <- read_csv("https://bit.ly/03-data-ca")
```

Load the remaining datasets.

```{r}
# Acadia National Park
acadia <- read_csv("https://bit.ly/03-data-acadia")

# Southeast US National Parks
se <- read_csv("https://bit.ly/03-data-se")

# 2016 Visitation for all Pacific West National Parks
visit_16 <- read_csv("https://bit.ly/03-data-visit-16")

# All Nationally designated sites in Massachusetts
mass <- read_csv("https://bit.ly/03-data-mass")
```






***

# Take aways

You can use this code template to make thousands of graphs with **ggplot2**.

```{r eval = FALSE}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```
