---
title: "Tidy"
output:
  pdf_document: default
  github_document: default
---

## Setup

```{r setup}
library(tidyverse)
```



## Warm up

Packages in the tidyverse include articles that explain how to use them. 

* Go to this specific of one tidyr's article. <https://tidyr.tidyverse.org/articles/pivot.html#pew>.
* Type the two lines of code you see in the second chunk.

```{r}
relig_income %>% 
  pivot_longer(-religion, names_to = "income", values_to = "count")
```

Explore the data, the article's section, and the code. Think:

* Why is the input messy and the output tidy?
* What does each argument do?



## Data

* Get the data.

```{r}
## wide messy format
gap_wide <- readr::read_csv('https://raw.githubusercontent.com/carpentries-incubator/open-science-with-r/gh-pages/data/gapminder_wide.csv')

## long tidy format
gapminder <- readr::read_csv('https://raw.githubusercontent.com/carpentries-incubator/open-science-with-r/gh-pages/data/gapminder.csv')
```

Bored?

* Create a new folder called "data"
* Download the data to your computer
* Upload it to rstudio.cloud into "data"
* Replace the URLs with paths to each file in your project.



Have a look at each dataset.

```{r}
gap_wide
gapminder
```

Bored?

* You know `head()` helps you see the first few rows. Which function helps you see the last few rows?
* Try `str()`. 
* Tweak `str()` with the arguments `list.len`, and `give.attr`



* From `gap_wide` select `country`, `continent` and columns starting with "gdp".

```{r}
gap_wide %>% 
  select(1, 2, starts_with("gdp"))
```



* Extend the previous pipeline.
* Pivot the dataset on the columns starting with "gdp" (the output is longer).

```{r}
gap_wide %>% 
  select(1, 2, starts_with("gdp")) %>%
  pivot_longer(cols = starts_with("gdp"))
```

Bored?

* Use arguments to get column names more informative than "name" and "value".


* Why is this longer?

```{r}
gw1 <- gap_wide %>% 
  pivot_longer(
    cols = starts_with("gdp") | starts_with("life") | starts_with("pop")
  )
gw1
```



* These alternatives are identical but shorter. How do they work?

```{r}
gw2 <- gap_wide %>% pivot_longer(cols = -continent:-country)
identical(gw1, gw2)

gw3 <- gap_wide %>% pivot_longer(cols = -1:-2)
identical(gw1, gw3)
```



`gw1` is still messy. Why?

* Tidy the dataset `gw1` using `separate()` and the vector c("metric", "year").

```{r}
tidy <- gw1 %>% separate(col = name, into = c("metric", "year"))
tidy
```

* Celebrate! With such a tidy dataset you are now ready to understand your data and communicate your work.
* Use the readr package to write your `tidy` dataset in "data/tidy.csv".
* You'll need to create the folder "data".

```{r}
if (!dir.exists("data")) {
  dir.create("data")
}

tidy %>% write_csv(path = "data/tidy.csv")
```
