---
title: "Tidy"
output: github_document
---

## Setup

```{r setup}
library(tidyverse)
```



## Warm up

This is a toy dataset. Why is it messy?

```{r}
cases <- tribble(
  ~Country, ~"2011", ~"2012", ~"2013",
      "FR",    7000,    6900,    7000,
      "DE",    5800,    6000,    6200,
      "US",   15000,   14000,   13000
)

cases
```



* Explain these two approaches to achieve the same tidy dataset.
* The result is tidy. Why?

```{r}
cases %>% 
  pivot_longer(cols = "2011":"2013")

cases %>% 
  pivot_longer(cols = -Country)
```



## Data

Read a messy dataset, then have a look.

```{r}
gap_wide <- readr::read_csv("data/gap_wide.csv")

gap_wide
```



* Pivot on the columns that start with "gdp" (see `?tidyselect::starts_with`).
* How long is the input?
* How long is the output?

```{r}
gap_wide %>% 
  pivot_longer(cols = starts_with("gdp"))
```



* Pivot on all columns except `continent` and `country`.
* Refer to the columns you want to pivot on using `starts_with()` (`|` means or)

```{r}
gap_wide %>% 
  pivot_longer(
    cols = starts_with("gdp") | starts_with("life") | starts_with("pop")
  )
```



This alternative is identical but shorter to type.

* The result is still messy. Why?

```{r}
gap_wide %>% 
  pivot_longer(cols = -continent:-country)
```



* Now also use `separate()` and the vector `c("metric", "year")`.
* We now achieved a tidy dataset. Why?

```{r}
tidy <- gap_wide %>% 
  pivot_longer(cols = -continent:-country) %>% 
  separate(col = name, into = c("metric", "year"))

tidy
```



* Use `unite()` and `pivot_wider()` to mess things up back again.

```{r}
messy <- tidy %>% 
  unite("name", metric, year) %>% 
  pivot_wider()

messy

gap_wide
```

## Filling missing data with `complete()`

This toy dataset has implicit missing data in year 2000. Why?

```{r}
kelp <- tibble(
  year = c(1999, 1999, 2000, 2004, 2004),
  taxon = c("Agarum", "Saccharina", "Saccharina", "Agarum", "Saccharina"),
  abundance = c(1, 4, 5, 8, 2)
)

kelp
```



* Use `complete()` to make explicit the missing data in `year` and `taxon`.

```{r}
kelp %>% 
  complete(year, taxon)
```



In `year = 2000` you found no individual of `taxon` "Agarum".

* Use the argument `fill` to fill missing data with the value `0`.

```{r}
kelp %>% 
  complete(year, taxon, fill = list(abundance = 0))
```



You surveyed every year between 1999-2004 but found nothing between 2001-2003.

```{r}
success_years <- kelp %>% 
  distinct(year) %>% 
  pull(year)

success_years
```

* Use `full_seq()` to produce the full sequence of the years you surveyed.

```{r}
all_years <- success_years %>% 
  full_seq(period = 1)

all_years
```



* Use `all_years` to `complete()` your dataset with explicit missing values.

```{r}
kelp %>% 
  complete(year = all_years, taxon = taxon)
```



You now want to study "Agarum" only.

```{r}
agarum <- kelp %>% 
  filter(taxon == "Agarum") %>% 
  complete(year = all_years)

agarum
```

* `fill()` the missing values of `taxon`.

```{r}
agarum %>% 
  fill(taxon)
```

* Now also `replace_na()` in `abundance` with the value `0`.

```{r}
agarum %>% 
  fill(taxon) %>% 
  replace_na(list(abundance = 0))
```



***

# Take Aways

Data comes in many formats but the tidyverse prefers just one: _tidy data_.

A data set is tidy if and only if:

1. Every variable is in its own column.
2. Every observation is in its own row.
