---
title: "Tidy"
output:
  github_document: default
  pdf_document: default
---

## Setup

```{r setup}
_______(tidyverse)
```



## Warm up

Packages in the tidyverse include articles that explain how to use them. 

* Go to this specific of one tidyr's article. <https://tidyr.tidyverse.org/articles/pivot.html#pew>.
* Type the two lines of code you see in the second chunk.

```{r}


```

Look and the data and code.

* Why is the input messy and the output tidy?
* Guess what each argument does.



## Data

* Read a messy dataset.

```{r}
gap_wide <- readr::________('https://raw.githubusercontent.com/carpentries-incubator/open-science-with-r/gh-pages/data/gapminder_wide.csv')
```



Have a look the dataset.

```{r}

```

Bored?

* You know `head()` helps you see the first few rows. Which function helps you see the last few rows?
* Try `str()`. 
* Tweak `str()` with the arguments `list.len`, and `give.attr`



* From `gap_wide` select `country`, `continent` and columns starting with "gdp".

```{r}
________ %>%
  ______(1, 2, starts_with("___"))
```



* Extend the previous pipeline.
* Pivot the dataset on the columns starting with "gdp" (the output is longer).

```{r}
________ %>% 
  ________________________________ %>%
  pivot_longer(cols = starts_with("___"))
```

Bored?

* Use arguments to get column names more informative than "name" and "value".



* Why is this longer?

```{r}
gw1 <- gap_wide %>% 
  pivot_longer(
    cols = starts_with("gdp") | starts_with("life") | starts_with("pop")
  )
gw1
```



* These alternatives are identical but shorter. How do they work?

```{r}
gw2 <- gap_wide %>% pivot_longer(cols = -continent:-country)
identical(gw1, gw2)

gw3 <- gap_wide %>% pivot_longer(cols = -1:-2)
identical(gw1, gw3)
```



`gw1` is still messy. Why?

* Tidy the dataset `gw1` using `separate()` and the vector c("metric", "year").

```{r}
tidy <- gw1 %>% ________(col = name, into = c("______", "____"))
tidy
```



* Use `unite()` and `pivot_wider()` to mess things up again.

```{r}
messy <- tidy %>% 
  _____("name", metric, year) %>% 
  ___________()

messy

gap_wide
```

