---
title: "Wrap up"
output: github_document
---

## Setup https://www.tidyverse.org/

```{r setup}
________(tidyverse)
```

## Import https://readr.tidyverse.org/

* Read messy data from <https://raw.githubusercontent.com/carpentries-incubator/open-science-with-r/gh-pages/data/gapminder_wide.csv>.

```{r}
url <- "https://raw.githubusercontent.com/carpentries-incubator/open-science-with-r/gh-pages/data/gapminder_wide.csv"

messy <- _______________(url)
messy
```

## Tidy https://tidyr.tidyverse.org/

* Pivot the data on everything except `continent` and `country`.
* Separate the messy column into `metric` and `year`.

```{r}
tidy <- messy %>% 
  ____________(cols = -continent:-country) %>% 
  # Watch out! year is a character vector, not a number
  separate(col = name, into = c("metric", "year"))

tidy
```



## Transform https://dplyr.tidyverse.org/

* Transform the year column from character to integer
* Filter it to include data since 1980.
* Select all columns except `country`.

```{r}
polished <- tidy %>% 
  ______(year = as.integer(year)) %>%
  ______(year > 1980) %>% 
  ______(-country)

polished
```

## Visualize https://ggplot2.tidyverse.org/

Plot a model for each metric through time:

* Map `color` and `fill` to `continent`.
* Plot model with uncertainty around the mean.
* Facet by `metric` (free the y variable -- see `scales`).

```{r}
ggplot(polished, aes(year, value, ____ = continent, _____ = continent)) +
  _____smooth() +
  facet_wrap(~ ______, ______ = "free_y")
```



## Communicate https://rmarkdown.rstudio.com/

* Create a report in .pdf and word format.

