---
title: "Wrap up"
output: github_document
---

<https://github.com/maurolepore/open-science-with-r>

**Before you start, collapse all chunks with Alt+o**



## Setup https://www.tidyverse.org/

* Use tidyverse packages.

```{r}
library(tidyverse)
```



## Import https://readr.tidyverse.org/

* Read messy data from "data/gap_wide.csv".

```{r}
messy <- readr::read_csv("data/gap_wide.csv")

messy
```



## Tidy https://tidyr.tidyverse.org/

* Pivot the data on everything except `continent` and `country`.
* Separate the messy column into `metric` and `year`.

```{r}
tidy <- messy %>% 
  pivot_longer(cols = -continent:-country) %>% 
  # Watch out! year is a character vector, not a number
  separate(col = name, into = c("metric", "year"))

tidy
```



## Transform https://dplyr.tidyverse.org/

* Transform the year column from character to integer
* Filter it to include data since 1980.
* Select all columns except `country`.

```{r}
polished <- tidy %>% 
  mutate(year = as.integer(year)) %>%
  filter(year > 1980) %>% 
  select(-country)

polished
```



## Visualize https://ggplot2.tidyverse.org/

Plot a model for each metric through time:

* Map `color` and `fill` to `continent`.
* Plot model with uncertainty around the mean.
* Facet by `metric` (free the `y` variable -- see `?scales`).

```{r}
ggplot(polished, aes(year, value, fill = continent, color = continent)) +
  geom_smooth() +
  facet_wrap(~ metric, scales = "free_y")
```



## Communicate https://rmarkdown.rstudio.com/

* Create a report in .md, .pdf, and word format.

